---
title: "Analyze TCGA data for SKCM cohort"
author: "Jacqueline Buros & ... "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here we are demonstrating the feasibility of analyzing genomic data using Stan. The first use case is to analyze somatic mutations for association with survival, after adjusting for key clinical variables with known prognostic status.

```{r setup, include = F, warnings = F, echo = F}

knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
# library(SuMu)
devtools::load_all('.')
library(dplyr)
library(survminer)
library(survival)
library(ggplot2)
library(scales)
```

## Clinical Data

First, download the clinical data. Here we are using the TCGA skin cutaneous melanoma (SKCM) cohort.

```{r load-clin}
clin_df <- SuMu::get_tcga_clinical(cohort = "SKCM")

## format some clinical data variables

clin_df2 <- clin_df %>% 
  dplyr::mutate(stage_part1 = gsub(pathologic_stage,
                                   pattern = '(Stage [0I]+).*',
                                   replacement = '\\1'),
                diagnosis_year_group = cut(year_of_initial_pathologic_diagnosis, breaks = c(1975, 1990, 1995, 2000, 2005, 2010, 2015, 2020), include.lowest = TRUE)
                ) %>%
  dplyr::rename(OS = `_OS`, OS_IND = `_OS_IND`)

```

### Review clinical data 

For this analysis we will consider the survival time in `r print(unique(clin_df['_OS_UNIT']))` since initial pathologic diagnosis.

```{r plot-survival}
fit <- survfit(Surv(OS, OS_IND) ~ 1,
                  data = clin_df2)
survminer::ggsurvplot(fit, ggtheme = theme_minimal())
```

Plotting by stage, although the time of 'stage' determination may be confounded if not collected at time of initial diagnosis.

```{r plot-survival-by-stage}
fit <- survfit(Surv(OS, OS_IND) ~ stage_part1,
                  data = clin_df2)
survminer::ggsurvplot(fit, ggtheme = theme_minimal(), legend = 'right')
```

There also seem to be differences by tumor type.

```{r plot-survival-by-sample-type}
fit <- survfit(Surv(OS, OS_IND) ~ sample_type,
                  data = clin_df2)
survminer::ggsurvplot(fit, legend = "right")
```

(Aside: I wonder how similar tumor type is to sample type? For example, we could have a metastatic patient where the sample was obtained from the primary tumor. We will want to adjust our genetic data analysis for the sample type but may want to estimate prognosis according to the tumor type?)

A variable like `year_of_initial_pathologic_diagnosis` is guaranteed to be unconfounded since we can safely assume it was collected at the time of diagnosis.

```{r plot-survival-by-year}
fit <- survfit(Surv(OS, OS_IND) ~ diagnosis_year_group,
                  data = clin_df2)
survminer::ggsurvplot(fit, ggtheme = theme_minimal(),
                      legend = 'right')
```

This makes it pretty clear that we have a strong "survival" bias to our data. This would suggest that, among people whose diagnosis was made in the 90s, only those who survived long enough to be enrolled were included in the study.

Let's look at a histogram of years of initial diagnosis.

```{r summary-init-diagnosis}
ggplot(clin_df2, aes(x = year_of_initial_pathologic_diagnosis,
                     fill = diagnosis_year_group)) + 
  geom_histogram() + 
  theme_minimal()
```

Let's look at the time since initial diagnosis (presumably, the time from enrollment to diagnosis).

```{r plot-time-since-diagnosis}

```

Finally, we can visualize a more comprehesive set of clinical variables.

```{r analyze-coxph}
fit <- survival::coxph(Surv(OS, OS_IND) ~
                         age_at_initial_pathologic_diagnosis + 
                         sample_type +
                         breslow_depth_value + initial_weight + 
                         strata(year_of_initial_pathologic_diagnosis),
               data = clin_df2)
print(fit)
```
