---
title: "Fitting a GLM model using Stan"
author: "Jacqueline Buros"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are two ways to fit a GLM model using [Stan](https://mc-stan.org).

1. Use rstanarm::stan_glm
2. Write the stan code for the model & fit using `rstan::fit`

We will work through both examples here.

## Using rstanarm

First let's select a dataset to which we want to fit our model.

```{r get-data}
data("mtcars")
```

Let's decide on the model we want to estimate. For now we will fit a really generic model to predict `mpg`.

```{r fit-model1}
library(rstanarm)
fit1 <- rstanarm::stan_glm(mpg / 10 ~ .,
                          data = mtcars)
print(fit1)
```

We can modify the priors on the covariates, to enforce greater sparsity.

```{r fit-model2}
fit2 <- rstanarm::stan_glm(mpg / 10 ~ .,
                          data = mtcars,
                          prior = cauchy(),
                          adapt_delta = 0.999
                          )
print(fit2)
```

We can also [put a prior on `R2`](https://rdrr.io/cran/rstanarm/f/vignettes/lm.Rmd). This reflects our overall confidence in the ability of the model to explain our outcome.

```{r fit-model3}
fit3 <- rstanarm::stan_lm(mpg / 10 ~ .,
                          data = mtcars,
                          prior = R2(location = 0.3, what = 'mode'),
                          adapt_delta = 0.999
                          )
print(fit3)
```


## evaluating posterior fits

Now, let's assume these all sampled well & that we have good estimates for our parameters from the three fits.

First thing we might want to do is compare these posterior estimates.

```{r compare-fits}
library(bayesplot)
plot1 <- bayesplot::mcmc_areas(as.array(fit1), facet_args = list(ncol = 1), pars = c('cyl', 'wt'))
plot2 <- bayesplot::mcmc_areas(as.array(fit2), facet_args = list(ncol = 1), pars = c('cyl', 'wt'))
plot3 <- bayesplot::mcmc_areas(as.array(fit3), facet_args = list(ncol = 1), pars = c('cyl', 'wt'))
gridExtra::grid.arrange(plot1, plot2, ncol = 2)
```
